Governing through patronage: the bargain for education in decentralized Brazil

### 2020-09-11 paper/paper.tex
redraft intro

In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to its capture by local political elites. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors ultimately cause turnover in the local educational bureaucracy, with negative downstream consequences for public school students.

To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage. Contrary to electoral accountability frameworks \citep{ferejohn_incumbent_1986,besley_political_2003}, mayors resort to patronage in order to shore up legislative support with limited consequences. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning. I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory. Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

>>>>

In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to its capture by local political elites. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

To analyze how politics affect staff turnover, I propose a theoretical model of legislative vote-buying, which I apply to the Brazilian context. I test the empirical implications of the model using Brazil as a case. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage. Contrary to electoral accountability frameworks \citep{ferejohn_incumbent_1986,besley_political_2003}, mayors resort to patronage in order to shore up legislative support with limited consequences. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning. I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory. Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.
